---
title: "World Cup 2026 Predictor — R + SQL"
author: "Tadaishe Maumbe"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

# Overview

This notebook is the R-language sibling of `python_sql_predictor.ipynb`. Both
read from the same SQLite database (`data/wc2026.sqlite`) built by
`src/build_db.py`, and both fit the same Dixon–Coles bivariate Poisson model
to international match data so the two language paths can be cross-checked.

The plan:

1. Connect to SQLite and inspect the three core tables (results, FIFA
   rankings, EA FC squad ratings).
2. Engineer a strength score per team that blends FIFA points and EA squad
   overall.
3. Fit Dixon–Coles on time-decayed historical international matches.
4. Layer the corners / cards baseline and the penalty shootout model.
5. Monte-Carlo the group stage (10 000 runs), resolve advancing teams,
   simulate the knockout bracket, write predictions back to the database.
6. Visualise the bracket and headline numbers.

```{r packages}
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(stringr)
library(lubridate)
library(MASS)
```

# 1. Connect to the database

```{r connect}
# Find the DB whether the notebook is run from notebooks/, the project root,
# or via rmarkdown::render() from anywhere else.
find_db <- function() {
  candidates <- c(
    file.path("..", "data", "wc2026.sqlite"),
    file.path("data", "wc2026.sqlite")
  )
  for (p in candidates) if (file.exists(p)) return(normalizePath(p))
  stop("Could not find wc2026.sqlite - run python src/build_db.py first")
}
db_path <- find_db()
con <- dbConnect(SQLite(), db_path)
dbListTables(con)
```

```{r inspect-tables}
tibble(
  table = c("historical_matches", "fifa_rankings", "player_ratings",
            "team_squad_strength", "wc2026_groups", "wc2026_fixtures"),
  rows  = sapply(c("historical_matches", "fifa_rankings", "player_ratings",
                   "team_squad_strength", "wc2026_groups", "wc2026_fixtures"),
                 function(t) dbGetQuery(con,
                   sprintf("SELECT COUNT(*) AS n FROM %s", t))$n)
)
```

# 2. Build the team strength feature

The strength score blends two signals:

- **EA FC 26 squad overall** (top-11 average) — reflects current talent.
- **FIFA ranking points** (most recent) — reflects recent results.

```{r strength}
sql_strength <- "
  WITH latest_rank AS (
    SELECT team, rank, points,
           ROW_NUMBER() OVER (PARTITION BY team ORDER BY rank_date DESC) AS rn
    FROM fifa_rankings
  )
  SELECT g.team,
         s.squad_overall_top11,
         s.squad_attack_mean,
         s.squad_defense_mean,
         r.rank   AS fifa_rank,
         r.points AS fifa_points
  FROM wc2026_groups g
  LEFT JOIN team_squad_strength s ON s.team = g.team
  LEFT JOIN latest_rank r ON r.team = g.team AND r.rn = 1;
"
strength <- dbGetQuery(con, sql_strength)

# Z-score blend
zscore <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
strength <- strength |>
  mutate(z_squad = zscore(squad_overall_top11),
         z_fifa  = zscore(fifa_points),
         strength = 0.6 * coalesce(z_squad, 0) + 0.4 * coalesce(z_fifa, 0))

strength |> arrange(desc(strength)) |> head(15)
```

# 3. Fit Dixon-Coles

```{r dc-data}
# Pull the last 6 years of matches. The Python notebook uses 2014+ but R's
# optim() is materially slower than scipy.optimize.minimize on the same
# problem - shorter horizon keeps the knit time tolerable while still
# covering two full World Cup cycles.
hist <- dbGetQuery(con, "
  SELECT match_date, home_team, away_team, home_score, away_score, neutral
  FROM historical_matches
  WHERE match_date >= '2020-01-01'
    AND home_score IS NOT NULL
    AND away_score IS NOT NULL;
") |>
  mutate(match_date = as.Date(match_date),
         neutral = as.integer(coalesce(neutral, 0)))

ref_date <- as.Date("2026-06-10")
hist <- hist |>
  mutate(days = as.numeric(ref_date - match_date),
         weight = exp(-0.0019 * days))
nrow(hist)
```

```{r dc-fit}
teams <- sort(unique(c(hist$home_team, hist$away_team)))
n <- length(teams)
idx <- setNames(seq_along(teams), teams)

dc_correction <- function(x, y, lam, mu, rho) {
  ifelse(x == 0 & y == 0, 1 - lam * mu * rho,
  ifelse(x == 0 & y == 1, 1 + lam * rho,
  ifelse(x == 1 & y == 0, 1 + mu * rho,
  ifelse(x == 1 & y == 1, 1 - rho, 1))))
}

neg_ll <- function(par) {
  attack  <- par[1:n] - mean(par[1:n])
  defense <- par[(n + 1):(2 * n)]
  ha <- par[2 * n + 1]
  rho <- par[2 * n + 2]
  hi <- idx[hist$home_team]; ai <- idx[hist$away_team]
  lam <- exp(attack[hi] + defense[ai] + ha * (1 - hist$neutral))
  mu  <- exp(attack[ai] + defense[hi])
  ll <- hist$home_score * log(lam) - lam +
        hist$away_score * log(mu)  - mu -
        lfactorial(hist$home_score) - lfactorial(hist$away_score)
  corr <- pmax(dc_correction(hist$home_score, hist$away_score, lam, mu, rho), 1e-12)
  -sum((ll + log(corr)) * hist$weight)
}

start <- c(rep(0, n), rep(0, n), 0.25, -0.05)
fit <- optim(start, neg_ll, method = "L-BFGS-B",
             control = list(maxit = 200))

attack  <- fit$par[1:n]; attack <- attack - mean(attack)
defense <- fit$par[(n + 1):(2 * n)]
home_adv <- fit$par[2 * n + 1]
rho      <- fit$par[2 * n + 2]
list(home_adv = home_adv, rho = rho, n_teams = n)
```

```{r dc-strengths}
dc_params <- tibble(team = teams, attack = attack, defense = defense) |>
  mutate(attack_rank = rank(-attack))
dc_params |> arrange(desc(attack)) |> head(15)
```

# 4. Score matrix helper

```{r score-matrix}
score_matrix <- function(home, away, neutral = TRUE, max_goals = 8) {
  if (!(home %in% teams) || !(away %in% teams)) return(NULL)
  ha <- if (neutral) 0 else home_adv
  lam <- exp(attack[idx[home]] + defense[idx[away]] + ha)
  mu  <- exp(attack[idx[away]] + defense[idx[home]])
  m <- outer(dpois(0:max_goals, lam), dpois(0:max_goals, mu))
  for (h in 0:1) for (a in 0:1) m[h + 1, a + 1] <-
    m[h + 1, a + 1] * dc_correction(h, a, lam, mu, rho)
  m / sum(m)
}

modal_score <- function(home, away, neutral = TRUE) {
  m <- score_matrix(home, away, neutral)
  i <- which(m == max(m), arr.ind = TRUE)[1, ]
  list(home = i[1] - 1, away = i[2] - 1)
}

# Sanity check
modal_score("Brazil", "Scotland")
```

# 5. Corners / cards baseline + penalty model

```{r baselines}
wc_avg <- list(
  GROUP = list(corners = 9.3, yellows = 3.8, p_red = 0.06),
  R32   = list(corners = 9.7, yellows = 4.5, p_red = 0.10),
  R16   = list(corners = 9.8, yellows = 4.7, p_red = 0.11),
  QF    = list(corners = 10.1, yellows = 5.0, p_red = 0.13),
  SF    = list(corners = 10.0, yellows = 5.2, p_red = 0.13),
  `3RD` = list(corners = 9.8, yellows = 4.6, p_red = 0.09),
  FINAL = list(corners = 10.3, yellows = 5.4, p_red = 0.14)
)

corners_baseline <- function(stage, gap) {
  total <- wc_avg[[stage]]$corners
  share <- pmin(0.8, pmax(0.2, 0.5 + 0.06 * gap))
  c(home = total * share, away = total * (1 - share))
}

cards_baseline <- function(stage, gap) {
  yel <- wc_avg[[stage]]$yellows
  pr  <- wc_avg[[stage]]$p_red
  w_share <- pmin(0.65, pmax(0.35, 0.5 + 0.04 * (-gap)))
  home_yel <- if (gap > 0) yel * (1 - w_share) else yel * w_share
  c(home_yel = home_yel, away_yel = yel - home_yel,
    p_home_red = max(0, min(1, pr / 2 * (1 + 0.1 * (-gap)))),
    p_away_red = max(0, min(1, pr / 2 * (1 + 0.1 * gap))))
}

shootout_rate <- c(Germany = 0.82, Argentina = 0.73, Brazil = 0.50,
  France = 0.58, Spain = 0.50, England = 0.30, Portugal = 0.50,
  Croatia = 0.75, Netherlands = 0.40, Belgium = 0.50, Mexico = 0.43,
  Morocco = 0.75, Senegal = 0.67, Uruguay = 0.55, `USA` = 0.60)

shootout_prob <- function(home, away) {
  h <- if (!is.na(shootout_rate[home])) shootout_rate[home] else 0.5
  a <- if (!is.na(shootout_rate[away])) shootout_rate[away] else 0.5
  unname(h / (h + a))
}
```

# 6. Single-match simulator

```{r sim-match}
simulate_match <- function(home, away, stage = "GROUP", neutral = TRUE,
                           knockout = FALSE) {
  m <- score_matrix(home, away, neutral = neutral); if (is.null(m)) return(NULL)
  flat <- as.vector(m); flat <- flat / sum(flat)
  k <- sample.int(length(flat), 1, prob = flat)
  max_g <- nrow(m) - 1
  h <- (k - 1) %/% (max_g + 1); a <- (k - 1) %% (max_g + 1)

  gap <- (attack[idx[home]] - attack[idx[away]]) +
         (defense[idx[away]] - defense[idx[home]])
  cb <- corners_baseline(stage, gap); cd <- cards_baseline(stage, gap)

  out <- list(
    home_score = h, away_score = a,
    home_corners = rpois(1, cb["home"]), away_corners = rpois(1, cb["away"]),
    home_yellows = rpois(1, cd["home_yel"]),
    away_yellows = rpois(1, cd["away_yel"]),
    home_red = as.integer(runif(1) < cd["p_home_red"]),
    away_red = as.integer(runif(1) < cd["p_away_red"]),
    went_to_pens = FALSE, home_advances = NA
  )
  if (knockout) {
    if (h == a) {
      out$went_to_pens <- TRUE
      out$home_advances <- runif(1) < shootout_prob(home, away)
    } else {
      out$home_advances <- h > a
    }
  }
  out
}
```

# 7. Monte Carlo the group stage

```{r mc-groups}
groups <- dbGetQuery(con, "SELECT group_id, team FROM wc2026_groups
                            ORDER BY group_id, team")
fixtures_grp <- dbGetQuery(con,
  "SELECT fixture_id, group_id, home_team, away_team
   FROM wc2026_fixtures WHERE stage = 'GROUP'")

N_SIM <- 10000
set.seed(2026)

simulate_group_stage <- function() {
  res <- fixtures_grp |>
    mutate(sim = pmap(list(home_team, away_team),
                      ~ simulate_match(..1, ..2, "GROUP")))
  res$h <- sapply(res$sim, `[[`, "home_score")
  res$a <- sapply(res$sim, `[[`, "away_score")
  res
}

standings_from <- function(rows) {
  long <- bind_rows(
    rows |> transmute(group_id, team = home_team, gf = h, ga = a),
    rows |> transmute(group_id, team = away_team, gf = a, ga = h)
  ) |>
    mutate(pts = if_else(gf > ga, 3L, if_else(gf == ga, 1L, 0L)))

  long |>
    group_by(group_id, team) |>
    summarise(pts = sum(pts), gd = sum(gf - ga), gf = sum(gf),
              .groups = "drop") |>
    group_by(group_id) |>
    arrange(desc(pts), desc(gd), desc(gf)) |>
    mutate(rank = row_number()) |>
    ungroup()
}

# Run once for illustration; the full 10k loop runs in the simulation chunk
one_run <- simulate_group_stage()
standings_from(one_run) |> filter(group_id == "L")
```

# 8. Knockout bracket simulation

For the 2026 format the top 2 of each group (24 teams) plus the 8 best
third-placed teams (32 total) enter the Round of 32. Below is a simplified
seeding: top seeds drawn against best 3rds, runners-up drawn against each
other. A production version would encode FIFA's official bracket map.

```{r knockout}
knockout_run <- function(advancers) {
  # advancers: tibble with team, seed (1=group winner, 2=runner-up, 3=best-3rd)
  bracket <- advancers |> arrange(seed, group_id)
  rounds <- list(R32 = bracket)
  for (stg in c("R32", "R16", "QF", "SF", "FINAL")) {
    teams_in <- rounds[[stg]]$team
    pairs <- matrix(teams_in, ncol = 2, byrow = TRUE)
    winners <- apply(pairs, 1, function(p) {
      r <- simulate_match(p[1], p[2], stg, knockout = TRUE)
      if (r$home_advances) p[1] else p[2]
    })
    next_stg <- switch(stg, R32 = "R16", R16 = "QF",
                       QF = "SF", SF = "FINAL", FINAL = NULL)
    if (!is.null(next_stg)) rounds[[next_stg]] <- tibble(team = winners)
  }
  rounds
}
```

# 9. Run the full Monte Carlo (10 000 tournaments)

Run this chunk with `eval = TRUE` when you are ready to commit ~5 minutes
of CPU. The cached results live in `tournament_sim`.

```{r mc-full, eval=FALSE}
champion_count <- integer()
for (i in seq_len(N_SIM)) {
  one <- simulate_group_stage()
  st <- standings_from(one)
  top2 <- st |> filter(rank <= 2)
  thirds <- st |> filter(rank == 3) |> arrange(desc(pts), desc(gd)) |>
    head(8)
  advancers <- bind_rows(
    top2 |> mutate(seed = rank),
    thirds |> mutate(seed = 3)
  )
  rounds <- knockout_run(advancers)
  champ <- rounds$FINAL$team[1]
  champion_count[champ] <- (champion_count[champ] %||% 0L) + 1L
}
tibble(team = names(champion_count),
       p_champion = champion_count / N_SIM) |>
  arrange(desc(p_champion)) |> head(15)
```

# 10. Write predictions back to the database

```{r write-preds, eval=FALSE}
preds <- fixtures_grp |>
  mutate(sim = pmap(list(home_team, away_team),
                    ~ {
                      m <- score_matrix(..1, ..2)
                      i <- which(m == max(m), arr.ind = TRUE)[1, ]
                      list(mh = i[1] - 1, ma = i[2] - 1,
                           pw_h = sum(m[lower.tri(m)]),
                           pw_d = sum(diag(m)),
                           pw_a = sum(m[upper.tri(m)]))
                    }))
# ... build into the predictions table with source = 'r'
# dbWriteTable(con, "predictions", out_df, append = TRUE)
```

# 11. Close

```{r close}
dbDisconnect(con)
```
